---
title: "Shift Times, Exercise, and Weight Gain"
subtitle: "Examining the Effects of Metabolic Activity and Shift Times on Weight Gain with a Zero-Inflated Poisson Model"
author: "P. Hunt, B. Kill, and B. Winters"
date: "October 5, 2021"
output: 
  beamer_presentation:
    theme: "Madrid"
header-includes:
  - \usepackage{dcolumn}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=F, warning=F)
library(readxl)
library(ggplot2)
library(dagitty)
library(lavaan)
library(ggdag)
library(lme4)
library(pscl)
library(faraway)
library(tidyverse)
library(ggpubr)
library(MASS)
library(kableExtra)

cb_palette = c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

data <- read_xlsx("Practicum 1 Data.xlsx")
p1.Data <- data[c("Job", "dept", "shift", "gender", "Age", "height", "weightgain", "lbs_gained", "bweight", "BMI", "Vig ex Time", "Walk ex Time", "Mod ex time", "Total_Met_Min")]
names(p1.Data)[names(p1.Data)=="Vig ex Time"] <- "Vig_ex_Time"
names(p1.Data)[names(p1.Data)=="Walk ex Time"] <- "Walk_ex_Time"
names(p1.Data)[names(p1.Data)=="Mod ex time"] <- "Mod_ex_Time"

p1.Data$Job  <- as.factor(p1.Data$Job)
p1.Data$shift  <- factor(p1.Data$shift, levels = c("7am", "8am", "9am", "10am","11am","12pm","1pm","2pm", "other"))
p1.Data$oshift <- unclass(p1.Data$shift)
p1.Data$gender  <- as.factor(p1.Data$gender)
p1.Data$dept  <- as.factor(p1.Data$dept)
p1.Data$weightgain  <- as.factor(p1.Data$weightgain)

p1.Data$Total_Met_Min <- 8*p1.Data$Vig_ex_Time + 4*p1.Data$Mod_ex_Time + 3.3*p1.Data$Walk_ex_Time
p1.Data$lbs_gained[p1.Data$weightgain == "No"] <- 0
p1.Data <- p1.Data[is.na(p1.Data$lbs_gained) == FALSE,]
p1.Data$beg_weight <- p1.Data$bweight-p1.Data$lbs_gained

p1.Data <- p1.Data[-c(1,2)]
p1.Data1 <- na.omit(p1.Data)


is_outlier <- function(x) {
  return(x <= quantile(x, 0.25) - 1.5 * IQR(x) | x >= quantile(x, 0.75) + 1.5 * IQR(x))
}

p1.Data.out <- p1.Data1 %>%
                  group_by(shift) %>%
                  mutate(outlier = ifelse(is_outlier(lbs_gained), "Yes", "No"))

p1.Data2 <- p1.Data.out %>%
                filter(outlier == "No")

p1.Data1 %>%
  group_by(shift) %>%
  mutate(outlier = ifelse(is_outlier(lbs_gained), lbs_gained, as.numeric(NA))) %>%
  ggplot(., aes(x = shift, y = lbs_gained, color = gender)) +
    geom_boxplot() +
    geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.3) +
  scale_color_manual(values  =  cb_palette[c(2,3)]) +
  labs(title = "Side-by-Side Boxplot of Pounds Gained vs. Shift split by Gender",
       subtitle = "Outliers labeled by their Pounds Gained",
       x= "Shift",
       y= "Weight Gain")

p1.Data2 %>%
  group_by(shift) %>%
  mutate(outlier = ifelse(is_outlier(lbs_gained), lbs_gained, as.numeric(NA))) %>%
  ggplot(., aes(x = shift, y = lbs_gained, color = gender)) +
    geom_boxplot() +
    geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.3) +
  scale_color_manual(values  =  cb_palette[c(2,3)]) +
  labs(title = "Side-by-Side Boxplot of Pounds Gained vs. Shift split by Gender",
       subtitle = "Outliers labeled by their Pounds Gained",
       x= "Shift",
       y= "Weight Gain" )
zipmod.ord <- zeroinfl(round(lbs_gained,0)~oshift:log(Total_Met_Min+1)+oshift:beg_weight+oshift*gender+log(Total_Met_Min+1)*gender+beg_weight*gender+Age, data = p1.Data)
zipmod.ord2 <- zeroinfl(round(lbs_gained,0)~oshift:log(Total_Met_Min+1)+oshift:beg_weight+oshift*gender+log(Total_Met_Min+1)*gender+beg_weight*gender+Age,data = p1.Data2)
p1.Data3 <- filter(p1.Data2, lbs_gained > 0)
zipmod.ord3 <- glm(round(lbs_gained,0)~oshift:log(Total_Met_Min+1)+oshift:beg_weight+oshift*gender+log(Total_Met_Min+1)*gender+beg_weight*gender+Age, data = p1.Data3, family = "poisson")
```

## Research Questions

\begin{block}{Q1: What is the effect of exercise time on weight gain?}
\begin{itemize}

\item Overall effect is \alert{negative}\\
\item Negative effect is \alert{weaker} with later shifts

\end{itemize}
\end{block}

\begin{block}{Q2: What is the effect of shift times (earlier to later) on weight gain?}
\begin{itemize}

\item Overall effect is \alert{negative}\\
\item Negative effect is \alert{stronger} in men

\end{itemize}
\end{block}

## Data

\begin{alertblock}{Response: Weight Gained}
\begin{itemize}
\item Treat as a count of pounds gained over an 8 month period
\item Distinguish between those who gained weight and those who did not
\end{itemize}
\end{alertblock}

```{r, echo=F, warning = F, message = F, fig.height= 2.5, fig.width=5}
set.seed(691)
jitter     <- runif(length(p1.Data$lbs_gained[p1.Data$weightgain=="Yes"]), -2.5, 2.5)
lbs_jitter <- p1.Data$lbs_gained
lbs_jitter[p1.Data$weightgain=="Yes"] <- lbs_jitter[p1.Data$weightgain=="Yes"] + jitter
  
ggplot(data.frame(lbs_jitter), aes(x=lbs_jitter,y= ..density..))+
  geom_histogram(color = cb_palette[1], fill = cb_palette[2]) +
  ggtitle("Histogram of Pounds Gained") + 
  xlab("Pounds Gained") + ylab("Frequency") +   
  geom_density(data = p1.Data[p1.Data$weightgain == "Yes",], aes(x=lbs_gained)) +
  labs(caption = "We can model the number of pounds gained as a Poisson 
       random variable with an inflated zero count.")
```

## Predictor Variables
\begin{block}{Key Predictor 1: Total Metabolic Minutes}
\begin{itemize}
\item A measure of weekly exercise time
\item A weighted sum of walk, moderate, and vigorous exercise
\end{itemize}
\end{block}

\begin{block}{Key Predictor 2: Shift Time}
\begin{itemize}
\item The time when the respondent's shift begins
\item Treated as ordinal
\end{itemize}
\end{block}

\begin{block}{Additional Control Variables}
\begin{itemize}
\item Beginning Weight (in lbs.)
\item Gender (Male/Female)
\item Age (in years)
\end{itemize}
\end{block}


## Missing Data and Outliers
\begin{alertblock}{Missing Observations}
\begin{itemize}
\item 392 responses from a population of 1,100 employees
\item 44 missing weight gain
\item 238 remaining after imputing TMM
\end{itemize}
\end{alertblock}

\begin{block}{Outliers}
\begin{itemize}
\item 8 observations of weight gain outside of 1.5*IQR range
\item Model fit improved with omission of these 8 observations
\begin{enumerate}
\item Log-likelihood rises from -1115 to -767.7
\item Diagnostic plots are improved
\end{enumerate}
\end{itemize}
\end{block}

## Model Building
### Causal Elicitation
- Causal questions require explicit modeling of causation
- Use caution in interpretation (this was not a RCT)

### Interactions
- No a priori theoretical expectation for interaction effects
- Checked all predictors for interactions with gender and shift
- Included only if different factor levels had non-parallel slopes


## Our Model
```{r, echo=FALSE, message=FALSE, warning=FALSE}
CI.graph <- dagitty('dag {
Job             [pos="0,1"]
Dept.           [pos="1,1"]
WET         [pos="3,1"]
MET          [pos="4,1"]
VET          [pos="5,1"]
Shift           [pos="2,2"]
TMM             [pos="4,2"]
Age             [pos="4,3"]
Gender          [pos="8,3"]
BW              [pos="6,3"]
WG     [pos="3,4"]

Job -> Dept. -> Shift
Job -> Shift -> TMM
Shift -> WG 
WET -> TMM
MET -> TMM
VET -> TMM -> WG
Age -> TMM
Age -> WG
Age -> BW
Gender -> TMM
Gender -> WG
Gender -> BW
BW -> WG
BW -> TMM
}')

ggdag(CI.graph) + theme_dag()+geom_dag_node(col = "gray", size=20) + geom_dag_text(col = "blue")+geom_dag_edges() + ggtitle("DAG of Causal Model")
```

## Interactions
### with Shift
- Total Metabolic Minutes
- Age
- Beginning Weight
- Gender

### with Gender
- Beginning Weight
- Shift

##  Transformations

Only TMM required a transformation (natural log) to reduce the skewdness.
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=7}
par(mfrow=c(2,1))
gg <- ggplot(group_by(p1.Data2, gender), aes(x=Total_Met_Min, y=..density..))+
  geom_histogram(color = cb_palette[1], fill = cb_palette[2]) + 
  geom_density() +
  scale_color_manual(values  =  cb_palette) +
  labs(title = "Histograms of Total Metabolic Minutes",
       x = "TMM") +
  facet_grid(.~gender)
gg1 <- ggplot(p1.Data2, aes(x=log(Total_Met_Min+1), y=..density..))+
  geom_histogram(color = cb_palette[1], fill = cb_palette[2]) + 
  geom_density(show.legend = FALSE) +
  scale_color_manual(values  =  cb_palette) +
  labs(x = "Log of TMM" ) +
  facet_grid(.~gender)

ggarrange(gg, gg1, nrow = 2)
```

## Final Regression Results
::: {.block}
### Zero-Inflated Poisson Model
- Binomial model for gaining weight or not:
    + Coefficients represent the change in log-odds of gaining weight
    + No significant predictors found in this part of the model

- Poisson Model for number of pounds gained:
    + Coefficients represent change in log-mean pounds gained
    + 6 significant predictors found in this part of the model
:::

Variable                        Coefficient   p-value
--------                        -----------   -------
Shift                           -0.356        <0.001
Log of Total Met Min            -0.145        <0.001
Beginning weight                \ 0.004       \ \ 0.036
Shift:log of Total Met Min      \ 0.055       <0.001
Shift:Male                      -0.133        <0.001
Beginning Weight:Male           \ 0.004       \ \ 0.007 

## Poisson Diagnostic Plots
- Variance = Mean
- Half-Normal Quantiles
- Null Residuals
```{r, echo=FALSE, message=FALSE, fig.height =2, fig.width=6}
par(mfrow = c(1,3))
plot(log(fitted(zipmod.ord3)), log((p1.Data3$lbs_gained-fitted(zipmod.ord3))^2),
     xlab=expression(hat(mu)[i]),
     ylab=expression((y[i]-hat(mu)[i])^2),
     main = "Variance vs. Mean",
     ylim = c(-8,8)); abline(0,1, col = "blue")



halfnorm(residuals(zipmod.ord3), main = "QQ Plot")
abline(0,2, col = "blue")

plot(fitted(zipmod.ord3), resid(zipmod.ord3),
     ylab="Residuals",
     xlab="Fitted Values",
     main = "Residuals vs. Fitted",
     ylim = c(-10,10)); abline(0,0, col = "blue")

```

## Discussion
:::{.block}
### Main Effects on Weight Gain
- TMM has a \alert{negative} effect
- Later shifts have a \alert{negative} effect
- Beginning weight has a \alert{small positive} effect
:::
:::{.block}
### Interactions
- Effect of TMM is \alert{weaker} with later shifts
- Effect of shift is \alert{stronger} in men
- Higher beginning weight is associated with \alert{greater} weight gain in men
:::
\begin{alertblock}{Caveats and Future Research}
\begin{itemize}
\item Need more research (preferably RCT) on who gains weight
\item Causal inference based on strong assumptions
\begin{itemize}
\item Should be tested further
\end{itemize}
\item Study does not distinguish between healthy and unhealthy weight gain
\end{itemize}
\end{alertblock}

## Appendex
Supplemental Tables and Figures.

## Poisson Results with Outliers
```{r, fig.width=4}
summ1 <- summary(zipmod.ord)
summ1$coefficients[[1]][,c(1,2,4)] %>%
  kbl(digits = 3, booktabs = T) %>%
  kable_classic(html_font = "Times New Roman")
```

## Zero-Inflated Estimates with Outliers 
```{r}
summ1$coefficients[[2]][,c(1,2,4)] %>%
  kbl(digits = 3, booktabs = T) %>%
  kable_classic(html_font = "Times New Roman")
```


## Poisson Estimates without Outliers
```{r, fig.width=4}
summ2 <- summary(zipmod.ord2)
summ2$coefficients[[1]][,c(1,2,4)] %>%
  kbl(digits = 3, booktabs = T) %>%
  kable_classic(html_font = "Times New Roman")
```

## Zero-Inflated Estimates without Outliers
```{r}
summ2$coefficients[[2]][,c(1,2,4)] %>%
  kbl(digits = 3, booktabs = T) %>%
  kable_classic(html_font = "Times New Roman")
```

## Shift by Gender
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=7}
ggplot(group_by(p1.Data2, gender), aes(x=shift, color = shift)) +
  geom_bar(color = cb_palette[1], fill = cb_palette[2]) + 
  scale_color_manual(values  =  cb_palette) +
  labs(title = "Barplot of Shift",
       subtitle = ,
       caption = ) +
  facet_grid(.~gender)
```

## Age and Log-transformation
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=7}
gg <- ggplot(group_by(p1.Data2, gender), aes(x=Age, y = ..density..)) +
  geom_histogram(color = cb_palette[1], fill = cb_palette[2], binwidth = 4) + 
  geom_density() +
  scale_color_manual(values  =  cb_palette) +
  labs(title = "a. Histogram of Age",
       x = "Age") +
  facet_grid(.~gender)

gg1 <- ggplot(group_by(p1.Data2, gender), aes(x = log(Age), y = ..density..)) +
  geom_histogram(color = cb_palette[1], fill = cb_palette[2], bins = 12) + 
  geom_density() +
  scale_color_manual(values  =  cb_palette) +
  labs(title = "b. Histogram of log(Age)",
       x= "Log of Age") +
  facet_grid(.~gender)

ggarrange(gg, gg1, nrow = 2)
```

## Weight and Log-Transformation
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=7}
gg <- ggplot(group_by(p1.Data2, gender), aes(x=beg_weight, y = ..density..)) +
  geom_histogram(color = cb_palette[1], fill = cb_palette[2], bins = 12) + 
  geom_density() +
  scale_color_manual(values  =  cb_palette) +
  labs(title = "Histogram of Beginning Weight",
       x = "Beginning Weight") +
  facet_grid(.~gender)

gg1 <- ggplot(group_by(p1.Data2, gender), aes(x=log(beg_weight), y = ..density..)) +
  geom_histogram(color = cb_palette[1], fill = cb_palette[2], bins = 12) + 
  geom_density() +
  scale_color_manual(values  =  cb_palette) +
  labs(title = "Histogram of log(Beginning Weight)",
       x = "Log of Beginning Weight") +
  facet_grid(.~gender)

ggarrange(gg, gg1, nrow = 2)
```

## TMM and Shift Interaction
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=5}
ggplot(p1.Data2[!is.na(p1.Data2$shift),], aes(x=Total_Met_Min, y=lbs_gained, color=shift)) +geom_point()+
  geom_smooth(method="lm", se= F) +
  scale_color_manual(values  =  rep(cb_palette,2)) + 
  labs(title = "Weight Gain vs. Total Metabolic Minutes",
       subtitle = "Split by Shift",
       x = "TMM (log scale)",
       y = "Pounds Gained",
       caption = "Non-parallel lines suggest interaction present") +
  scale_x_log10()
```

## Age and Shift Interaction
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=5}
ggplot(p1.Data2[!is.na(p1.Data2$shift),], aes(x=Age, y=lbs_gained, color=shift)) +geom_point()+
  geom_smooth(method="lm", se= F) +
  scale_color_manual(values  =  rep(cb_palette,2)) +
  labs(title = "Weight Gain vs. Age",
       subtitle = "Split by Shift",
      x = "Age",
      y= "Pounds Gained",
      caption = "Non-parallel lines suggest interaction present")
```

## Beginning Weight and Shift Interaction
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=5}
ggplot(p1.Data2[!is.na(p1.Data2$shift),], aes(x=beg_weight, y=lbs_gained, color=shift)) +geom_point()+
  geom_smooth(method="lm", se= F) + 
  scale_color_manual(values  =  rep(cb_palette,2)) +
  labs(title = "Weight Gain vs. Beginning Weight",
       subtitle = "Split by Shift",
       x = "Beginning Weight",
       y= "Pounds Gained",
       caption = "Non-parallel lines suggest interaction present")
```

## Gender and Shift Interaction
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=5}
ggplot(p1.Data2[!is.na(p1.Data2$gender),], aes(x=oshift, y=lbs_gained, color=gender)) +geom_point()+
  geom_smooth(method="lm", se = FALSE) +
  scale_color_manual(values  =  rep(cb_palette,2)[2:3]) +
  labs(title = "Weight Gain vs. Shift",
       subtitle = "Split by Gender",
       y="Pounds Gained",
       x="Shift",
       caption = "Non-parallel lines suggest interaction present") +
  scale_x_continuous(breaks = seq(1,10,1), labels = c("1" = "7am","2" = "8am","3"= "9am","4"= "10am","5"= "11am","6"= "12pm","7"= "1pm","8"="2pm","9"= "other","10"= "NA"))
```


## TMM and Gender Interaction
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=5}
ggplot(p1.Data2[!is.na(p1.Data2$gender),], aes(x=Total_Met_Min, y=lbs_gained, color=gender)) +geom_point()+
  geom_smooth(method="lm", se = FALSE) +
  scale_color_manual(values  =  rep(cb_palette,2)[2:3]) +
  labs(title = "Weight Gain vs. Total Metabolic Minutes", 
       subtitle = "Split by Gender",
       x = "TMM (log scale)",
       y = "Pounds Gained",
       caption = "Parallel lines suggest no interaction present") +
  scale_x_log10()
```


## Age and Gender Interaction
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=5}
ggplot(p1.Data2[!is.na(p1.Data2$gender),], aes(x=Age, y=lbs_gained, color=gender)) +geom_point()+
  geom_smooth(method="lm", se = FALSE) +
  scale_color_manual(values  =  rep(cb_palette,2)[c(2,3)]) +
  labs(title = "Weight Gain vs. Age",
       subtitle = "Split by Gender",
       x = "Age",
       y = "Pounds Gained",
       caption = "Parallel lines suggest no interaction present")
```

## Beginning Weight and Gender Interaction
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=5}
ggplot(p1.Data[!is.na(p1.Data$gender),], aes(x=beg_weight, y=lbs_gained, color=gender)) +geom_point()+
  geom_smooth(method="lm", se = FALSE) +
  scale_color_manual(values  =  rep(cb_palette,2)[c(2,3)]) +
  labs(title = "Weight Gain vs Beginning Weight",
       subtitle = "Split by Gender",
       x = "Beginning Weight",
       y = "Pounds Gained",
       caption = "Non-parallel lines suggest interaction present")
```